***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**

### 2015-10-19 Schloss_Cluster_PeerJ_2015.Rmd
Editing

---
title: "***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**"
bibliography: references.bib
output:
  markdown
  pdf_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

**Results.** The stability of OTU assignments did not reflect the quality of the assignments. Depending on the dataset being analyzed, the average linkage and the distance and abundance-based greedy clustering methods generated more robust OTUs than the open and closed-reference methods. Further interrogation of the reference-based methods indicated that when USEARCH is used to identify the closest reference, the OTU assignments were sensitive to the order of the reference sequences. When USEARCH was replaced by VSEARCH the quality and stability of the OTU assignments by the reference-based methods were improved. Finally, we demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH.

**Discussion.** This study demonstrates that the quality of clustering assignments needs to be assessed for multiple methods per dataset to identify the optimal clustering method for the dataset. The analysis also calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current versions of QIIME.

@@

The first approach has been referred to as phylotyping [@Schloss2011] or closed reference clustering [@NavasMolina2013]. This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference does not adequately reflect the biodiversity of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Therefore, defining OTUs in the closed-reference approach is complicated because although two sequences might be 97% similar to the same reference sequence, they may only be 94% similar to each other. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011]. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.

The second approach has been referred to as distance-based [@Schloss2011] or *de novo* clustering [@NavasMolina2013]. In this method, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures then these problems can be overcome [@Kozich2013]. As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010, @Sun2009, @Mah2014]. One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015; @Hamady2009]. Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011]. For this reason, *de novo* clustering has been preferred across the field.

The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013]. Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. In theory, this method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions employed by both approaches poses a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold. Those sequences that did not classify to a known family would also be clustered using the average linkage method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.

>>>>

---
title: "***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**"
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

**Results.** The stability of OTU assignments did not reflect the quality of the assignments. Depending on the dataset being analyzed, the average linkage and the distance and abundance-based greedy clustering methods generated more robust OTUs than the open and closed-reference methods. We also demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH. Further interrogation of the reference-based methods indicated that when USEARCH is used to identify the closest reference, the OTU assignments were sensitive to the order of the reference sequences because the reference sequences can be identical over the region being considered. More troubling was the observation that while both USEARCH and VSEARCH have a high level of sensitivity to detect reference sequences, the specificity of those matches was poor relative to the true best match.

**Discussion.** This study demonstrates that the quality of clustering assignments needs to be assessed for multiple methods per dataset to identify the optimal clustering method for the dataset. The analysis also calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current version of QIIME.

@@

The first approach has been referred to as phylotyping [@Schloss2011] or closed reference clustering [@NavasMolina2013]. This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference does not adequately reflect the biodiversity of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene [@Schloss2011]. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Therefore, defining OTUs in the closed-reference approach is complicated because although two sequences might be 97% similar to the same reference sequence, they may only be 94% similar to each other. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011]. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.

The second approach has been referred to as distance-based [@Schloss2011] or *de novo* clustering [@NavasMolina2013]. In this method, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures, then these problems can be overcome [@Kozich2013]. As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010, @Sun2009, @Mah2014]. One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015; @Hamady2009]. Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011]. For this reason, *de novo* clustering has been preferred across the field.

The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013; @Rideout2014]. Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. In theory, this method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions employed by both approaches poses a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold. Those sequences that did not classify to a known family would also be clustered using the average linkage method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.
